---
title: 'Class 10: Genome Informatics'
author: "Soobin (PID:A15201229)"
date: "2/17/2022"
output:
  pdf_document: default
  html_document: default
---

> Q5: What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)? 

```{r}
# Read genotype file from Ensembl

mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

table(mxl$Genotype..forward.strand.) / nrow(mxl)
```

What about a different population? Here we take the British in England and Scotland (GBR)

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

table(gbr$Genotype..forward.strand.) / nrow(gbr)
```

## Expression by Genotype Analysis

I want to read my RNA-Seq Expression results into R. This file is not a CSV but rather has fields separated by space. 

```{r}
expression <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expression)
```

Now we will look at the other genotypes.
```{r}
table(expression$geno)

```

```{r}
summary( expression[ expression$geno == "A/A", 3 ] )
summary( expression[ expression$geno == "A/G", 3 ] )
summary( expression[ expression$geno == "G/G", 3 ] )
```

## Make a summary overview figure

Make a boxplot figure
```{r}
library(ggplot2)
ggplot(expression) + aes(x=geno, y=exp, fill=geno) + geom_boxplot(notch=T) + labs(x="Genotype", y="Expression")
```

